\pdftitle

### 2017-05-17 maxNt.tex
new material on projection

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-17T10:31:16+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

By marginalization we obtain the probability for the state of the sample:
\begin{gather}
  \label{eq:marginal}
  \p(\yx = \yr \| \yH) = \binom{n}{n\yrs}^{-1} \p(\yxs=\yrs \| \yH),
  \\
  \shortintertext{with}
  \label{eq:subpop_average}
  \p(\yxs=\yrs \| \yH) = \sum_{N \yRf =0}^{N}
    \p(\yxs = \yrs \|\yXf=\yRf, \yH)\,
  \p(\yXf=\yRf \| \yH),
  \\[2\jot]
  \p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}.
\end{gather}

The latter conditional probability is a hypergeometric distribution,
typical of \enquote{drawing without replacement} problems. The
combinatorial proof of the formulae above is in fact the same as for this
class of problems
\cites[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.
This conditional probability acts a

The symmetry present in the probabilities above is not a physical property
of the neuronal population. It only expresses the symmetry of our initial
ignorance about the population, and does not imply any sort of physical
similarity between the neurons. Subsequent observations may in fact break
this symmetry. The initial symmetry should intuitively also apply to the
sample, and this is indeed the case: the probability for the state of the
sample~\eqref{eq:subpop_average} automatically satisfies the representation
theorem~\eqref{eq:joint_plaus_N_homog} as well.

\begin{figure}[!t]
\centering
%\includegraphics[width=0.99\linewidth]{pop_sample_projection.pdf}\\
\includegraphics[width=0.99\linewidth]{pop_sample_projection2.pdf}%
\caption{hyperg}
\label{fig:hypergeom_proj}
\end{figure}%hypergeometric_identity.nb
% \end{wrapfigure}
Our initial knowledge (ignorance) $\yH$ has two important properties
derivable from the formulae above.

First, our initial expectations for the averages of activity products
\emph{are the same for the sample and for the full population}:
\begin{equation}
  \label{eq:expe_products}
\expe{\sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}} \| \yH}
=
\expe{\av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}} \| \yH}
%\\[2\jot]
%{}
=
\binom{n}{m}^{-1}
%\frac{(n-m)!}{n!}
\sum_{n \yrs=0}^n %m!\, 
\binom{n \yrs}{m}\, \p(\yxs=\yrs \| \yH)
% \sum_{N \yRf=0}^N %m!\, 
% \binom{N-m}{N \yRf-m}
% \binom{N}{N \yRf}^{-1} \p(\yXf=\yRf \| \yH)
=
\binom{N}{m}^{-1}
%\frac{(N-m)!}{N!}
\sum_{N \yRf=0}^N %m!\, 
\binom{N \yRf}{m}\, \p(\yXf=\yRf \| \yH).
\end{equation}
Raw moments like $\expe{\yxs^m}$, which can be written in terms of the
product expectations above via \eqn~\eqref{eq:expe_products}, similarly
satisfy $\expe{\yxs^m} \to \expe{\yXf^m}$ as $n \to N$.

Second, upon observation of a sample average, say $\yxxs=a$, the updated
expectations for such average in the population \emph{and in any new
  sample} will usually be shifted towards the observed value, as follows
from Bayes's theorem and the formulae above.

These two properties express the fact that the sample is
\emph{representative} of the population. Our initial knowledge $\yH$ thus
assumes that by observing a sample we can make predictions about the full
population and about new samples. Contrast this with an assumption that
assigns an independent probability to each neuron: in this case, the
observation of one sample does not change our initial probability about
another -- we can't learn from observation.

@@

Formulae \eqref{eq:joint_plaus_N_homog}--\eqref{eq:expe_products} are
constraints on our initial probability assignment, but do not determine it
numerically. The probability $\p(\yXf=\yRf \| \yH)$ for the population
average needs to be numerically specified, and by \eqref{eq:subpop_average}
it will determine that of the sample average, $\p(\yxs=\yrs \| \yH)$. If we
numerically specify the latter, the former is not completely specified,
because \eqn~\eqref{eq:subpop_average} linearly constrains $N+1$ unknowns
by only $n+1$ equations in this case.

We may want to specify the probability by enforcing several of its sample
expectations to have specific values, for example $\expe{\yxs}=c_1$,
$\expe{\yxxs}=c_2$. This is still an underdetermined linear problem:
 several distributions can have the same desired expectations, as
clear from \eqns~\eqref{eq:expe_products}.

The \me\ method is brought into play to solve this indeterminacy. It selects
one distribution, purported to be \enquote{maximally noncommittal}, among
those that have the desired expectations. But we are in a quandary:
formulae~\eqref{eq:expe_products} allow us to apply the method to find the
probability of the population $\p(\yXf=\yRf \| \yH)$, or of the sample
$\p(\yxs=\yrs \| \yH)$. \emph{The two applications, however, are
  inequivalent}. They lead to numerically different $\p(\yxs=\yrs \| \yH)$.
Here is an example.

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-17T14:42:01+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

By marginalization we obtain the probability for the state of the sample:
\begin{gather}
  \label{eq:marginal}
  \p(\yx = \yr \| \yH) = \binom{n}{n\yrs}^{-1} \p(\yxs=\yrs \| \yH),
  \\
  \shortintertext{with}
  \label{eq:subpop_average}
  \p(\yxs=\yrs \| \yH) = \sum_{N \yRf =0}^{N}
    \p(\yxs = \yrs \|\yXf=\yRf, \yH)\,
  \p(\yXf=\yRf \| \yH),
  \\[2\jot]
  \label{eq:conditional_hypergeometric}
  \p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}
  \defs \ypp(\yrs\|\yRf).
\end{gather}

The latter conditional probability is a hypergeometric distribution
$\ypp(\yrs\|\yRf)$, typical of \enquote{drawing without replacement}
problems. The combinatorial proof of the formulae above is in fact the same
as for this class of problems
\cites[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.

The conditional probability $\ypp(\yrs,\yRf)$ connects the spaces of the
sample average $\yXf\in\set{0,\dotsc,N}$ and of the population average
$\yxs\in\set{0,\dotsc,n}$. It is a projector $\ypp$ of any probability $p$
for $\yXf$ onto a marginal probability $p_*$ for $\yxs$:
\begin{equation}
  \label{eq:projection_P}
  % \ypp\colon p(\yXf=\dotv) \to
  p_*(\yxs=\yrs)
  = \sum_{N\yRf=0}^N \ypp(\yrs\|\yRf) p(\yXf=\yRf).
\end{equation}
Therefore it also a pulls back expectations of functions $f$ of the sample
average $\yxs$ to expectations of functions $f^*$ the population average
$\yXf$:
\begin{equation}
  \label{eq:pullback_P}
%  \ypp^*\colon f(\yxs=\dotv) \to f^*(\yXf=\dotv) \defd \sum_{N\yrs=0}^n
%  f(\yrs)\ypp(\yrs\|\dotv),
  \begin{gathered}
    f^*(\yXf) \defd \sum_{n\yrs=0}^n f(\yrs)\ypp(\yrs\|\yXf),
    \\
   \expeb{f(\yxs)}= \expeb{f^*(\yXf)}
    =
    \sum_{n\yrs=0}^n f(\yrs)\,\p(\yxs=\yrs \| \yH)
    =
    \sum_{N\yRf=0}^N f^*(\yRf)\,\p(\yXf=\yRf \| \yH).
  \end{gathered}
\end{equation}

\begin{figure}[!t]
\centering
%\includegraphics[width=0.75\linewidth]{pop_sample_projection.pdf}\\
\includegraphics[width=0.75\linewidth]{pop_sample_projection2.pdf}%
\caption{Log-plot of the hypergeometric distribution
  $%\p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1} \defs
  \ypp(\yrs\|\yRf)$ for $N=5000$, $n=200$. (Band artifacts may appear in the
  colourbar depending on your \textsc{pdf} viewer.)}
\label{fig:hypergeom_proj}
\end{figure}%hypergeometric_identity.nb
A look at a plot of the hypergeometric distribution $\ypp(\yrs\|\yRf)$, for
example \fig~\ref{fig:hypergeom_proj}, reveals that it is a sort of
\enquote{fuzzy identity matrix} between the $\yXf$-space $\set{0,\dotsc,N}$
and $\yxs$-space $\set{0,\dotsc,n}$. When $n=N$ it is the identity matrix.
We thus have that, very roughly,
\begin{equation}
  \label{eq:roughly_equal_nN}
  \p(\yxs=a) \sim \p(\yXf=a),\qquad
\expeb{f(\yxs)} \sim \expeb{f(\yXf)}.
\end{equation}
These formulae express the fact that \emph{our uncertainty about the sample
  is representative of our uncertainty about the population and about other
  samples}, and vice versa. Upon observation of a sample average, say
$\yxxs=a$, the updated expectations for such average in the population and
in any new sample will usually be shifted towards the observed value, as
follows from Bayes's theorem and the formulae above.

Note that the approximate formulae~\eqref{eq:roughly_equal_nN} say more
than the limits $\p(\yxs=a) \to \p(\yXf=a)$,
$\expeb{f(\yxs)} \to \expeb{f(\yXf)}$ as $n\to N$. These limits are
trivially valid because the sample becomes the full population as $n\to N$.
In particular, these limits hold even in cases where the conditional
probability $\p(\yxs = \yrs \|\yXf=\yRf)$ is not a fuzzy identity and our
uncertainties about sample and about population can differ wildly.

The symmetry present in the probabilities above is not a physical property
of the neuronal population. It only expresses the symmetry of our initial
uncertainty about the population, and does not imply any sort of physical
similarity between the neurons. Subsequent observations may in fact break
this symmetry. The initial symmetry should intuitively also apply to the
sample, and this is indeed the case: the probability for the state of the
sample~\eqref{eq:subpop_average} automatically satisfies the representation
theorem~\eqref{eq:joint_plaus_N_homog} as well.

% The projection~\eqref{eq:projection_P} and pull-back~\eqref{eq:pullback_P}
% have the property
% \begin{equation}
%   \label{eq:expectation_pullback}
%   \expeb{f(\yxs) \|\yH} = \expeb{f^*(\yXf)\|\yH}
%   \qquad\text{or}\qquad
%   \sum_{n\yrs=0}^n f(\yrs)\,p_*(\yxs=\yrs) = \sum_{N\yRf=0}^N f^*(\yRf)\,p(\yXf=\yRf).
% \end{equation}
For functions representing averaged products,
$f(\yxs) \defd \sav{\yx\dots\yx}=\binom{\yxs}{m}/\binom{n}{m}$,
formulae~\eqref{eq:pullback_P} have the useful form
\begin{equation}
  \label{eq:expe_products}
%  \label{eq:pullback_m_expectations}
%  \ypp^*\colon \sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}}
%  \mapsto
  \begin{gathered}
    \bigl( \sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}}\bigr)^* =
    \av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}},
    \\[2\jot]
    % \expe{\sav{\underbrace{\yx \dotsm \yx}_{\text{$m$ factors}}} \| \yH} =
    % \expe{\av{\underbrace{\yX \dotsm \yX}_{\text{$m$ factors}}} \| \yH}
    \expe{\yx \dotsm \yx \| \yH} =
    \expe{\yX \dotsm \yX \| \yH}
    % \\[2\jot]
    % {}
    = \binom{n}{m}^{-1}
    % \frac{(n-m)!}{n!}
    \sum_{n \yrs=0}^n %m!\,
    \binom{n \yrs}{m}\, \p(\yxs=\yrs \| \yH)
    % \sum_{N \yRf=0}^N %m!\,
    % \binom{N-m}{N \yRf-m} \binom{N}{N \yRf}^{-1} \p(\yXf=\yRf \| \yH)
    = \binom{N}{m}^{-1}
    % \frac{(N-m)!}{N!}
    \sum_{N \yRf=0}^N %m!\,
    \binom{N \yRf}{m}\, \p(\yXf=\yRf \| \yH).
  \end{gathered}
\end{equation}
The proof uses the expression for the $m$th factorial moment of the
hypergeometric distribution \cite{potts1953}. Thus, the averages of
activity products \emph{are the same for the sample and for the full
  population}. Similar relations can be found for the raw moments
$\expe{\yxs^m}$ and $\expe{\yXf^m}$, which can be written in terms of the
product expectations via \eqn~\eqref{eq:products_intermsof_average}.

@@

Formulae
\eqref{eq:joint_plaus_N_homog}--~\eqref{eq:conditional_hypergeometric} are
constraints on our initial probability assignment, but do not determine it
numerically. The probability $\p(\yXf=\yRf \| \yH)$ for the population
average needs to be numerically specified, and by \eqref{eq:subpop_average}
it will determine that of the sample average, $\p(\yxs=\yrs \| \yH)$. If we
numerically specify the latter, the former is not completely specified,
because \eqn~\eqref{eq:subpop_average} linearly constrains $N+1$ unknowns
by only $n+1$ equations in this case.

We may want to specify the probability by enforcing the sample expectations
of several functions to have specific values, for example
$\expe{\yxs}=c_1$, $\expe{\yxxs}=c_2$. This is still an underdetermined
linear problem: several distributions can have the same desired
expectations, as clear from \eqns~\eqref{eq:expe_products}.

The \me\ method is brought into play to solve this indeterminacy. It selects
one distribution, purported to be \enquote{maximally noncommittal}, among
those that have the desired expectations. But we are in a quandary:
formulae~\eqref{eq:pullback_P} allow us to apply the method to find the
probability of the population $\p(\yXf=\yRf \| \yH)$, or of the sample
$\p(\yxs=\yrs \| \yH)$. \emph{The two applications, however, are
  inequivalent}. They lead to numerically different $\p(\yxs=\yrs \| \yH)$.
Here is an example.
